Fused regression for network inference in multiple species

### 2015-05-21 paper1.tex

\begin{abstract}
Learning gene regulatory networks is essential to our understanding of cell behavior and response to external stimuli. Methods for network inference have been developed and used for a wide variety of species; however, these approaches consider the problem as independent in each species even though there is known to be conservation of gene regulation even in distantly related species. We introduce a method that allows data to be pooled across species in a principled way based on gene orthology and demonstrate that it improves network inference even when the orthology is incomplete. We further refine this method with an algorithm that attempts to learn the true conserved subnetwork from a larger set of potentially conserved interactions. 
\end{abstract}

@@

\section{Methods}
\subsection{Gene regulatory network}
We represent gene regulatory networks as a matrix of weights, where rows index transcription factors and columns index genes. The weight in a given position represents the regulatory network on that TF on that gene; positive weights represent activation, negative weights represent repression, and 0 weights represent the absence of an interaction. We begin with an existing framework for estimating the interaction weights in a gene regulatory network (GRN). In the Inferelator algorithm, linear differential equations are used to model changes in gene expression. The rate at which the observed mRNA expression of gene $i$, $x_i$, changes is governed by degradation of existing transcripts with rate $\alpha$ plus a linear combination of transcription factor (TF) expressions. 
\begin{equation}
\frac{\mathrm d}{\mathrm d t} x_i = -\alpha_{i}x_{i} + \sum \beta_{i,j}x_{j}
\end{equation}
We are interested in learning $\beta$, the matrix defining the influence of each TF on each gene. We fix the decay rate $\alpha$ for all genes, and set it assuming a time-constant of 10 minutes [CITATIONS], as in [CITATIONS]. Let $x_i(t)$ be the expression of gene $i$ at time $t$. Given time-series data on the expression of gene $i$ at timepoints $t_k$ and $t_{k+1}$, we can approximate the rate of change of $x_i$ as $x_i'(t_k)=\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k}$. We can treat steady-state data as having a derivative of zero. This gives us, for each gene $i$ and time $t_{k}$ an equation

\begin{equation}
\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k} + \alpha_{i}x_{i}(t_k)= \sum \beta_{i,j}x_{j}(t_k)
\end{equation}
\noindent We can summarize these equations in matrix form as
\begin{equation}
Y = \beta X
\end{equation}
\noindent We approach the learning of the $\beta$ matrix using regression. Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is essentially the same basic formulation as was used in the Inferelator algorithm, which we extend to the case of simultaneously inferring the GRNs of multiple species.

@@

\subsection{Solving}
The problem of solving a multi-valued regression problem can be converted into the problem of solving a single valued regression problem by vectorization.
Specifically, if $Y \in R^{N \times M}$ then we form $A =$ diagonally concatenate $X$, $M$ times (with the remaining entries 0),
$y =$ vertically concatenate the columns of $Y$

@@

\subsection{Optimization}
We use cross-validation to select optimal regularization parameters $\lambda_R$ and $\lambda_S$. In order to determine the relevant scale of the fusion weight $\lambda_S$, we can compute the correlation between pairs of coefficients linked by fusion constraints. The maximal value of $\lambda_S$ should be just large enough so that this correlation is near $1$. More details on this later.

>>>>

\begin{abstract}
Learning gene regulatory networks is essential to our understanding of cell behavior and response to external stimuli. Methods for network inference from large datasets have been developed and used for a wide variety of species; however, these approaches consider the problem as independent in each species even though there is known to be conservation of gene regulation even in distantly related species. We introduce a method that allows data to be pooled across species in a principled way based on gene orthology and demonstrate that it improves network inference even when the orthology is incomplete. We further refine this method with an algorithm that attempts to learn the true conserved subnetwork from a larger set of potentially conserved interactions. 
\end{abstract}

@@

\section{Methods}
\subsection{Gene regulatory network}
We represent gene regulatory networks as a matrix of weights, where rows index transcription factors and columns index genes. The weight in a given position represents the regulatory weight of that TF on that gene; positive weights represent activation, negative weights represent repression, and 0 weights represent the absence of an interaction. We begin with an existing framework for estimating the interaction weights in a gene regulatory network (GRN). In the Inferelator algorithm, linear differential equations are used to model changes in gene expression. The rate at which the observed mRNA expression of gene $i$, $x_i$, changes is governed by degradation of existing transcripts with rate $\alpha$ plus a linear combination of transcription factor (TF) expressions. 
\begin{equation}
\frac{\mathrm d}{\mathrm d t} x_i = -\alpha_{i}x_{i} + \sum \beta_{i,j}x_{j}
\end{equation}
We are interested in learning $\beta$, the matrix defining the influence of each TF on each gene. We fix the decay rate $\alpha$ for all genes, and set it assuming a time-constant of 10 minutes [CITATIONS], as in [CITATIONS]. Let $x_i(t)$ be the expression of gene $i$ at time $t$. Given time-series data on the expression of gene $i$ at timepoints $t_k$ and $t_{k+1}$, we can approximate the rate of change of $x_i$ as $x_i'(t_k)=\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k}$. We can treat steady-state data as having a derivative of zero. This gives us, for each gene $i$ and time $t_{k}$ an equation

\begin{equation}
\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k} + \alpha_{i}x_{i}(t_k)= \sum \beta_{i,j}x_{j}(t_k)
\end{equation}
\noindent We can summarize these equations in matrix form as
\begin{equation}
Y = X \beta 
\end{equation}
\noindent We approach the learning of the $\beta$ matrix using regression. Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is essentially the same basic formulation as was used in the Inferelator algorithm, which we extend to the case of simultaneously inferring the GRNs of multiple species.

@@

\subsection{Solving using augmented matrices}
The problem of solving a multi-valued regression problem can be converted into the problem of solving a single valued regression problem by vectorization.
Specifically, if $Y \in R^{N \times M}$ then we form $A =$ diagonally concatenate $X$, $M$ times (with the remaining entries 0),
$y =$ vertically concatenate the columns of $Y$

@@

\subsection{Optimization}
We use cross-validation to select optimal regularization parameters $\lambda_R$ and $\lambda_S$. In order to determine the relevant scale of the fusion weight $\lambda_S$, we can compute the correlation between pairs of coefficients linked by fusion constraints. The maximal value of $\lambda_S$ should be just large enough so that this correlation is near $1$.
